What We Eval in the Shadows

### 2021-08-10 paper/main.tex
Fix DSL description

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\mypara}[1]{\medskip\noindent\emph{#1}\xspace}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
\newcommand{\code}[1]{\lstinline |#1|\xspace}
%\renewcommand{\c}[1]{\lstinline |#1|\xspace}
% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother
\lstdefinelanguage{smalleR} {
  morekeywords={
    for,
    if,
    else,
    function
  },
  sensitive=true, % keywords are not case-sensitive
  morecomment=[l]{\#}, % l is for line comment
  morestring=[b]{"} % defines that strings are enclosed in double quotes
}
\lstset{
  language={smalleR},
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
%  framexleftmargin=1mm,
%  framexrightmargin=1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  alsoletter={., _, $},
  breaklines=true
}

@@

\begin{abstract}
  \noindent Most dynamic languages allow users to turn text into code using
  various functions, often named \eval, with language-dependent semantics. The
  widespread use of these reflective functions hinders static analysis and
  prevents compilers from performing optimizations. This paper aims to provide a
  better sense of why programmers use \eval. Understanding why \eval is used in
  practice is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and application
  domain-specific; we focus on data science code written in R and compare our
  results to previous work that analyzed web programming in JavaScript. This
  paper studied \CranRunnableScripts scripts extracted from \CranPackages R
  packages, for a total of \packageAllcalls calls to \eval. We find that \eval
  is indeed in widespread use; R's \eval is more pervasive and arguably
  dangerous than what was previously reported for JavaScript.
\end{abstract}

@@

\noindent The expression passed to \eval can be thought of as an abstract syntax
tree. It can be parsed from a string, created manually or using reflection.

@@

\noindent The call to \c{substitute(x)} extracts the unevaluated expression from
the promise \c{x}. The function \c{match.call} returns an expression
representing the current call.

R is generous when it comes to what is allowed to be passed as environment to
\eval. Besides ordinary environments, it also accepts a list or a data
frame (using element names or column names for variable resolution) or an
integer \c{n} (in which case it would use the \c{n.} call frame). If no
environment is specified, then the default is the environment where the call to
\eval was made (as returned by \c{parent.frame}).
%
Environments nest, each has a parent. When creating a new environment with
\c{new.env}, the parent is the current environment. Environment chains can be
traversed with \c{parent.env}, until \c{emptyenv} is reached. The top-level
environment is \c{.GlobalEnv}, it has parents that represent the packages that
have been loaded. One can also directly read, modify or create new bindings,
given any environment:
%
\begin{lstlisting}
  # reading                                # writing
  envir$v                                  envir$v <- 2
  get("v", envir=envir)                    assign("v", 2, envir=envir)
\end{lstlisting}
%
\noindent Environments are used as hash maps as they have reference semantics
and a built-in string lookup.

@@

\mypara{Discussion.} When we started this project, our goal was to contrast the
usage of \eval by package developers and end-users, but the Kaggle dataset have
so few calls to \eval that there is little that can be said about those. The
remainder of the paper focuses on CRAN with a few observations about the other
sources when relevant.

@@

       \node [block, fill = red, draw = darkred, very thick] (download) {
         \nodename{
           \vspace{1mm}\textbf{Download}\vspace{1mm}\\
           15K packages\\
           8K Kaggle kernels\\
           219 Kaggle datasets
         }
       };
       \node [below = \textsep of download]   (downloaddesc)   {\nodedesc{12 hours}};
       \node [block, right = \nodesep of download, fill = yellow, draw = darkyellow, very thick] (extract) {
         \nodename{
           \vspace{1mm}\textbf{Extract}\vspace{1mm}\\
           240K package programs\\%, 4.6M LOC\\
           74 Kaggle kernels\\%, 13K LOC\\
           \newline
         }
       };
       \node [below = \textsep of extract]   (extractdesc)   {\nodedesc{2 hours}};
       \node [block, right = \nodesep of extract, fill = blue, draw = darkblue, very thick] (trace) {
         \nodename{
           \vspace{1mm}\textbf{Trace}\vspace{1mm}\\
           49M eval calls \\
           14M side-effects\\
           \newline
         }
       };
       \node [below = \textsep of trace]     (tracedesc)     {\nodedesc{36 hours}};
       \node [block, right = \nodesep of trace, fill = green, draw = darkgreen, very thick] (analyze) {
         \nodename{
           \vspace{1mm}\textbf{Analyze}\vspace{1mm}\\
           2.8M unique eval calls\\
           22\% packages use eval\\
           \newline
         }
       };
       \node [below = \textsep of analyze]    (analyzedesc)    {\nodedesc{2 hours}};
       \draw [connector] (download)   edge (extract);
       \draw [connector] (extract)    edge (trace);
       \draw [connector] (trace)      edge (analyze);
     \end{tikzpicture}
   }
   \vspace{-2mm}\caption{Pipeline}\label{fig:pipeline}
 \end{figure}

@@

\noindent The pipeline runs in parallel~\cite{GNUparallel} orchestrated by a
Makefile. Servers have identical environments thanks to a docker image with all
dependencies installed.

@@

\mypara{Limitations.} Even with the extension described above, there are
\PkgUndefinedRnd \eval calls without source information (\PkgUndefinedRatio of
all \eval calls). This happen when \eval is an argument to a higher-order
function or when it is called from native code. An other limitation is that we
do not record calls to the native \eval. Neither of these limitations should
invalidate our conclusions.

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Measuring Eval across Datasets}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{wrapfigure}{r}{5cm} \vspace{-1cm} \hspace*{-12mm}
  \centering
  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf}
  \caption{CRAN \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}\vspace{-3mm}
\end{wrapfigure}
%
This section focuses on CRAN packages and reports statistics about the usage of
\eval. We use the word \emph{site} to refer to an occurrence of a call site to the \eval
function in the source code and \emph{call} to denote an observed invocation of
the \eval function.

\begin{figure}[!b]
\small
\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
\begin{minipage} {5cm}\small
  \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
    \multicolumn{3}{|c}{\small\#calls} &\small \#pck
&     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
\tt 1 &--& \tt 10      & \packageBina  & \tt 10K &--&\tt 100K  & \packageBine\\
\tt 11 &--& \tt 100    & \packageBinb  & \tt 100K &--&\tt 1M  & \packageBinf\\
\tt 101 &--& \tt 1K    & \packageBinc  & \tt 1M &--&\tt 10M   & \packageBing\\
\tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\hline
\end{tabular}
\caption{Call frequency}\label{freq}
\end{minipage}
&
\begin{minipage}{7cm}\small
\begin{tabular}{|@{\,}r|rrrr|}\hline
  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-1.5mm]
           & & & \c{.parent} &\\\hline
\small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
\small Exercised sites&\packageTriggeredeval&\packageTriggeredevalq&\packageTriggeredevalparent&\packageTriggeredlocal\\
\small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\hline
\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
\end{minipage}\end{tabular}

\end{figure}

\begin{figure}[b] \centering
  \includegraphics[width=.7\textwidth]{traced-eval-callsites.pdf} \centering
  \caption{\eval call sites coverage of the \PkgPackages packages.}%
  \label{fig:traced-eval-callsites}
\end{figure}

@@

  \vspace*{-1mm}
  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{package_calls_per_run_per_call_site}

@@

\caption{Normalized calls} \label{cn}\vspace{-2mm}

\medskip
\medskip

\begin{tabular}{c}
  \vspace*{-1mm}
  {\hspace{-25mm}\includegraphics[width=0.7\textwidth]{package_size_loaded_distribution}}
\end{tabular}
\caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}
There are \PkgEvalCallSites \eval sites in \PkgPackages packages. The proportion
of packages calling \eval is \PkgPackagesRatio. Over half of these packages
have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage, which
has \MaxEvalCallSitesCount sites, all packages contain fewer than
\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}, which
shows a histogram of sites per package). These sites appear in \PkgFunsWithEval
functions (\CranFunsWithEvalRatio of all functions in CRAN).
For dynamic analysis, we run \CranRunnableScripts programs extracted from
\CranPackages packages. Any run that does not exercise \eval is discarded. This
left \packageNbruns runs from \packageCorpus packages. There were
\packageAllcalls calls in \packageTriggeredpkgs packages originating from
\PkgHitEvalCallSites unique sites. In terms of coverage, the data exercised
\PkgHitEvalCallSitesAvgRatio of sites, a coverage similar to the code coverage
metric for the packages, which is \PkgCodeCoverage. The fact that not all sites
are exercised can be chalked down to incomplete tests and occasional analysis
failures (\PkgFailedProgramsRatio of programs crashed or timed out).
Fig~\ref{fig:traced-eval-callsites} shows the number of sites that were
exercised; coverage is unequal. Figure~\ref{freq} summarizes the frequency of
dynamic calls to \eval, with the left
column being the number of calls and the right, the number of packages in that
range. There are \packageFewcalls packages with low \eval frequency, fewer than
100 calls, and a smaller, but still significant, number of packages,
\packageManycalls to be precise, that use \eval more than 1,000 times. Package
\packageMaxcallspack makes \packageMaxcalls calls and thus accounts for over
half of the observations. Figure~\ref{tab:variantseval} summarizes the use of
variants of \eval. For each of the four variants, the first row shows the number
of sites in the corpus (\emph{static}), the second row is the number of sites
that were encountered during analysis (\emph{exercised}), the last row is the
number of calls (\emph{invocations}). The overwhelming majority of sites and
calls are to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
\c{local} are barely used at all. The difference between sites and exercised
sites underscores the limits of code coverage.

Figure~\ref{cn}(a) shows normalized call counts per site; on the left are the
average number of calls from a given site and given run, on the right are the
counts of sites that fall in that range. For instance, \packageRunbinh sites are
invoked 2,000+ times per run. Larger numbers suggest loops or recursive contexts
-- this seems to be the exception as most \evals, \packageRunbina, are exercised
50 times or less. Figure~\ref{cn}(b) zooms in on low frequency sites. The x-axis
shows normalized calls, and the y-axis is the number of sites for that value.
Most low-frequency \evals are invoked only once; about half as many are invoked
twice; after that, the frequency quickly drops.

\Eval takes any value, but if its argument is not an expression, it is returned
unchanged. Expressions account for \packageCodepercent of arguments. More
specifically, \packageSymbolpercent are \texttt{symbol}s (single variables such
as \c{x}), \packageLanguagepercent are \c{language} objects (function calls),
  and \packageExpressionpercent are \c{expression} objects (lists of
  expressions). Further inspection reveals that most symbols,
  \packageGgplotsymbolpercent to be exact, come from a single site in the
  \c{ggplot2} package and have the value \c{\_inherit}.\footnote{This models
  inheritance in \c{ggproto}, one of the many object-oriented systems in R, used
  in the \c{ggplot2} graphics library.}

@@

\mypara{Discussion.}
\Eval is used in CRAN packages. Nearly every fourth package contains at least
one \eval site. However, most packages that use \eval do that modestly. Half of
them include less than three \eval sites. Packages that rely heavily on \eval
are mostly the ones that provide statistical modeling and simulation functions.
At runtime, over a third of extracted programs triggered calls to \eval. Half
of the \packageAllcallsRnd recorded calls come from a single \eval site, but
for the rest, most of the packages have low \eval call frequency.

@@

\medskip\noindent  We detail these forms and discuss their implication for the behavior of \eval.

@@

\medskip\noindent\EE{$min(e)=\c{V}$} Expressions that represent values are
frequently passed to \eval -- they occur in 17\% of sites. The majority of
those, \packageValOneNodePercent, are inline constants (integer or double
vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
long as base functions such as \c{+} are not redefined. They typically are not,
but this is a limitation of this categorization.} In our corpus,
\packageNbCallSitesUniqueActualValue call sites only ever see a simple value,
such as:
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
where the argument to \eval is a string which \eval simply returns. Of twenty
randomly selected \evals, XXXXX
%% UNFINED TODO \packageUsefulValueEvalPercent
of them need \eval for
some inputs. Either we are dealing with a value that needs to be constructed
dynamically, or the value is a default case that sometimes is replaced by a more
interesting expression. This form is usually evaluated in few interpreter steps;
in fact, the median is only \packageMinimizedmedianoperationscRnd. The
environment in which they evaluate is mostly irrelevant.

\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form;
they are found in 28\% of the sites. This form includes simple variable reads,
\eg~\c{x}, which are \packageNbSymbolVarSitePercent of \c X. The form also
subsumes \c{V}, so it includes a mixture of arithmetic expressions,
\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
be bound to a promise, and accessing it may trigger evaluation of that promise,
thus resulting in an arbitrary amount of computation. The median number of
interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
it is not the common case. Lookups are often evaluated in constructed
environments; as few as \packageMinimizedpercentparentframesa of these
expressions are evaluated in a function environment.

\medskip\noindent\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
\c{x[[24]]}. As with \c X, we allow arithmetics and values in this form. Lookup
occurs in 10\% of the sites. The interpreter evaluates
\packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
operations. This is typically used in a function environment,
\packageMinimizedpercentparentframese of the time to be precise.

\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignment
operators, the direct assignment {\tt <-}, and assignment to the parent
environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form.
Assignments occur in 5\% of the sites. They represent the most obvious source of
side effects. The median count is \packageMinimizedmedianoperationsiRnd (min=3).

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables nor assignments. More specifically, it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then the computation will occur. This form occurs in
6\% of sites and typically does not perform much work.

\medskip\noindent\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
forms allow for function calls whose arguments may include variable references.
The latter allows nested calls. Assignments are excluded from this form. They
occur in, respectively, 14\% and 20\% of the sites in the corpus. Together they
are the most frequent forms. The median numbers of interpreter steps are,
respectively, \packageMinimizedmedianoperationsdRnd and
\packageMinimizedmedianoperationsbRnd. Also,
\packageMinimizedpercentparentframesd and \packageMinimizedpercentparentframesb
of these expressions run in function environments.

\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that
define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs
in \packageFunctionDefinitionSitesPercent of sites. Evaluating a function
definition is done in 2 interpreter steps; the data does not record the work
performed by the interpreter when the generated functions are eventually run. In
addition to \c{FUN}, \packageGeneralizedFunctionDefinitionSitesPercent of sites
have function definitions nested in other expressions. This gives an idea of the
use of higher-order functions.

\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
code blocks, which occur in only 4\% of sites. These are huge expressions; we
do not inspect the contents of the blocks. The median number of executed operations
is \packageMinimizedmedianoperationsjRnd. They typically run in function
environments.

>>>>

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
\newcommand{\code}[1]{\lstinline |#1|\xspace}
%\renewcommand{\c}[1]{\lstinline |#1|\xspace}
% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother
\lstdefinelanguage{smalleR} {
  morekeywords={
    for,
    if,
    else,
    function
  },
  sensitive=true, % keywords are not case-sensitive
  morecomment=[l]{\#}, % l is for line comment
  morestring=[b]{"} % defines that strings are enclosed in double quotes
}
\lstset{
  language={smalleR},
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
%  framexleftmargin=1mm,
%  framexrightmargin=1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  alsoletter={., _, $},
  breaklines=true
}

@@

\begin{abstract}
 Most dynamic languages allow users to turn text into code using
  various functions, often named \eval, with language-dependent semantics. The
  widespread use of these reflective functions hinders static analysis and
  prevents compilers from performing optimizations. This paper aims to provide a
  better sense of why programmers use \eval. Understanding why \eval is used in
  practice is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and application
  domain-specific; we focus on data science code written in R and compare our
  results to previous work that analyzed web programming in JavaScript. This
  paper studied \CranRunnableScripts scripts extracted from \CranPackages R
  packages, for a total of \packageAllcalls calls to \eval. We find that \eval
  is indeed in widespread use; R's \eval is more pervasive and arguably
  dangerous than what was previously reported for JavaScript.
\end{abstract}

@@

The expression passed to \eval can be thought of as an abstract syntax
tree. It can be parsed from a string, created manually or using reflection.

@@

 The call to \c{substitute(x)} extracts the unevaluated expression from
the promise \c{x}. The function \c{match.call} returns an expression
representing the current call.

R is generous when it comes to what is allowed to be passed as environment to
\eval. Besides ordinary environments, it also accepts a list or a data
frame (using element names or column names for variable resolution) or an
integer \c{n} (in which case it would use the \c{n.} call frame). If no
environment is specified, then the default is the environment where the call to
\eval was made (as returned by \c{parent.frame}).
%
Environments nest, each has a parent. When creating a new environment with
\c{new.env}, the parent is the current environment. Environment chains can be
traversed with \c{parent.env}, until \c{emptyenv} is reached. The top-level
environment is \c{.GlobalEnv}, it has parents that represent the packages that
have been loaded. One can also directly read, modify or create new bindings,
given any environment:
%
\begin{lstlisting}
  # reading                                # writing
  envir$v                                  envir$v <- 2
  get("v", envir=envir)                    assign("v", 2, envir=envir)
\end{lstlisting}
%
 Environments are used as hash maps as they have reference semantics
and a built-in string lookup.

@@

\paragraph{Discussion.} When we started this project, our goal was to contrast the
usage of \eval by package developers and end-users, but the Kaggle dataset have
so few calls to \eval that there is little that can be said about those. The
remainder of the paper focuses on CRAN with a few observations about the other
sources when relevant.

@@

       \node [block, fill = red, draw = darkred, very thick] (download) {
         \nodename{
           \vspace{1mm}\textbf{Download}\vspace{1mm}\\
           15K packages\\
           8K Kaggle kernels\\
           219 Kaggle datasets
         }
       };
       \node [below = \textsep of download]   (downloaddesc)   {\nodedesc{12 hours}};
       \node [block, right = \nodesep of download, fill = yellow, draw = darkyellow, very thick] (extract) {
         \nodename{
           \vspace{1mm}\textbf{Extract}\vspace{1mm}\\
           240K package programs\\%, 4.6M LOC\\
           74 Kaggle kernels\\%, 13K LOC\\
           \newline
         }
       };
       \node [below = \textsep of extract]   (extractdesc)   {\nodedesc{2 hours}};
       \node [block, right = \nodesep of extract, fill = blue, draw = darkblue, very thick] (trace) {
         \nodename{
           \vspace{1mm}\textbf{Trace}\vspace{1mm}\\
           49M eval calls \\
           14M side-effects\\
           \newline
         }
       };
       \node [below = \textsep of trace]     (tracedesc)     {\nodedesc{36 hours}};
       \node [block, right = \nodesep of trace, fill = green, draw = darkgreen, very thick] (analyze) {
         \nodename{
           \vspace{1mm}\textbf{Analyze}\vspace{1mm}\\
           2.8M unique eval calls\\
           22\% packages use eval\\
           \newline
         }
       };
       \node [below = \textsep of analyze]    (analyzedesc)    {\nodedesc{2 hours}};
       \draw [connector] (download)   edge (extract);
       \draw [connector] (extract)    edge (trace);
       \draw [connector] (trace)      edge (analyze);
     \end{tikzpicture}
   }
   \caption{Pipeline}\label{fig:pipeline}
 \end{figure}

@@

 The pipeline runs in parallel~\cite{GNUparallel} orchestrated by a
Makefile. Servers have identical environments thanks to a docker image with all
dependencies installed.

@@

\paragraph{Limitations.} Even with the extension described above, there are
\PkgUndefinedRnd \eval calls without source information (\PkgUndefinedRatio of
all \eval calls). This happen when \eval is an argument to a higher-order
function or when it is called from native code. An other limitation is that we
do not record calls to the native \eval. Neither of these limitations should
invalidate our conclusions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Measuring Eval}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section reports on the frequency of \eval in CRAN. We use \emph{site} to
refer to an occurrence of a call site to the \eval function in the source code
and \emph{call} to denote an observed invocation of the \eval function.

There are \PkgEvalCallSites \eval sites in \PkgPackages packages;
\PkgPackagesRatio packages use \eval. Over half of these packages have fewer
than 3 sites, and with the exception \MaxEvalCallSitesPackage which has
\MaxEvalCallSitesCount sites, all packages contain fewer than
\MaxEvalCallSitesRest sites. Fig~\ref{fig:pkgs-eval-callsites-hist} shows a
histogram of sites per package. Sites appear in \PkgFunsWithEval functions;
\CranFunsWithEvalRatio of all functions use \eval.

\begin{figure}[!b]
\centering
  \includegraphics[width=10cm]{pkgs-eval-callsites-hist.pdf}
  \caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}

  \medskip

  \small
\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
\begin{minipage} {5cm}\small
  \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
    \multicolumn{3}{|c}{\small\#calls} &\small \#pck
&     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
\tt 1 &--& \tt 10      & \packageBina  & \tt 10K &--&\tt 100K  & \packageBine\\
\tt 11 &--& \tt 100    & \packageBinb  & \tt 100K &--&\tt 1M  & \packageBinf\\
\tt 101 &--& \tt 1K    & \packageBinc  & \tt 1M &--&\tt 10M   & \packageBing\\
\tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\hline
\end{tabular}
\caption{Call frequency}\label{freq}
\end{minipage}
&
\begin{minipage}{7cm}\small
\begin{tabular}{|@{\,}r|rrrr|}\hline
  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-1.5mm]
           & & & \c{.parent} &\\\hline
\small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
\small Exercised sites&\packageTriggeredeval&\packageTriggeredevalq&\packageTriggeredevalparent&\packageTriggeredlocal\\
\small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\hline
\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
\end{minipage}\end{tabular}

\medskip

\includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf} \centering
  \caption{\eval call sites coverage of the \PkgPackages packages.}%
  \label{fig:traced-eval-callsites}
\end{figure}

@@

  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{package_calls_per_run_per_call_site}

@@

\caption{Normalized calls} \label{cn}

\medskip\medskip
\begin{tabular}{c}
{\hspace{-25mm}\includegraphics[width=0.7\textwidth]{package_size_loaded_distribution}}
\end{tabular}
\caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}

Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
packages. Any run that does not call \eval is discarded, leaving us with
\packageNbruns runs. We recorded \packageAllcalls \eval calls originating from
\PkgHitEvalCallSites sites. This exercised \PkgHitEvalCallSitesAvgRatio of
sites, the coverage is similar to the package code coverage metric which is
\PkgCodeCoverage. The reason some sites are not exercised can be chalked down to
incomplete tests and analysis failures (\PkgFailedProgramsRatio of all scripts
crashed or timed out). Fig~\ref{fig:traced-eval-callsites} plots exercised site
with respect to all sites for each program; as can be seen coverage is unequal.
Figure~\ref{freq} summarizes the frequency of calls, number of calls on the left
and number of packages on the right. There are \packageFewcalls packages with
fewer than 100 calls, and \packageManycalls packages with more than 1,000 calls.
\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
over half of the observations. Fig.~\ref{tab:variantseval} summarizes use of
variants, the first row has sites (\emph{static}), the second has sites
encountered during analysis (\emph{exercised}) and the last gives calls
(\emph{invocations}). Most sites and calls go to \eval itself, \c{eval.parent}
is rare, and both \c{evalq} and \c{local} are barely used at all.

Figure~\ref{cn}(a) shows average count of calls from a given site and given run
and how many sites fall in that range. For instance, \packageRunbinh sites are
called over 2,000 times per run. Larger numbers suggest loops or recursive
contexts -- this seems to be the exception; \packageRunbina of sites are called
fewer than 50 times. Fig.~\ref{cn}(b) zooms in on low frequency sites. The
x-axis shows normalized calls, and the y-axis counts sites. Most sites are
invoked once, half as many are invoked twice, the drop off is sharp after that.

\Eval accepts any value as argument, but if the value is not an expression,
\eval returns it unchanged. Expressions account for \packageCodepercent of
arguments. More specifically, \packageSymbolpercent are \c{symbol}s (variables
such as \c{x}), \packageLanguagepercent are \c{language} objects (expressions
representing a function calls), and \packageExpressionpercent are \c{expression}
objects (lists of expressions). Further inspection reveals that \c{\_inherit}
accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
\c{ggplot2}.\footnote{To model inheritance in \c{ggproto}, one of the many
object-oriented systems for R.}

@@

\paragraph{Discussion.}
\Eval is used in CRAN packages. Nearly every fourth package contains at least
one \eval site. However, most packages that use \eval do that modestly. Half of
them include less than three \eval sites. Packages that rely heavily on \eval
are mostly the ones that provide statistical modeling and simulation functions.
At runtime, over a third of extracted programs triggered calls to \eval. Half
of the \packageAllcallsRnd recorded calls come from a single \eval site, but
for the rest, most of the packages have low \eval call frequency.

@@

\medskip  We detail these forms and discuss their implication for the behavior of \eval.

@@

\medskip\EE{$min(e)=\c{V}$} Expressions that represent values are
frequently passed to \eval -- they occur in 17\% of sites. The majority of
those, \packageValOneNodePercent, are inline constants (integer or double
vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
long as base functions such as \c{+} are not redefined. They typically are not,
but this is a limitation of this categorization.} In our corpus,
\packageNbCallSitesUniqueActualValue call sites only ever see a simple value,
such as:
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
where the argument to \eval is a string which \eval simply returns. Of twenty
randomly selected \evals, XXXXX
%% UNFINED TODO \packageUsefulValueEvalPercent
of them need \eval for
some inputs. Either we are dealing with a value that needs to be constructed
dynamically, or the value is a default case that sometimes is replaced by a more
interesting expression. This form is usually evaluated in few interpreter steps;
in fact, the median is only \packageMinimizedmedianoperationscRnd. The
environment in which they evaluate is mostly irrelevant.

\medskip\EE{$min(e)=\c{X}$} Variable lookups are the most common form;
they are found in 28\% of the sites. This form includes simple variable reads,
\eg~\c{x}, which are \packageNbSymbolVarSitePercent of \c X. The form also
subsumes \c{V}, so it includes a mixture of arithmetic expressions,
\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
be bound to a promise, and accessing it may trigger evaluation of that promise,
thus resulting in an arbitrary amount of computation. The median number of
interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
it is not the common case. Lookups are often evaluated in constructed
environments; as few as \packageMinimizedpercentparentframesa of these
expressions are evaluated in a function environment.

\medskip\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
\c{x[[24]]}. As with \c X, we allow arithmetics and values in this form. Lookup
occurs in 10\% of the sites. The interpreter evaluates
\packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
operations. This is typically used in a function environment,
\packageMinimizedpercentparentframese of the time to be precise.

\medskip\EE{$min(e)=$~\c{<-}} This form includes both assignment
operators, the direct assignment {\tt <-}, and assignment to the parent
environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form.
Assignments occur in 5\% of the sites. They represent the most obvious source of
side effects. The median count is \packageMinimizedmedianoperationsiRnd (min=3).

\medskip\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables nor assignments. More specifically, it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then the computation will occur. This form occurs in
6\% of sites and typically does not perform much work.

\medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
forms allow for function calls whose arguments may include variable references.
The latter allows nested calls. Assignments are excluded from this form. They
occur in, respectively, 14\% and 20\% of the sites in the corpus. Together they
are the most frequent forms. The median numbers of interpreter steps are,
respectively, \packageMinimizedmedianoperationsdRnd and
\packageMinimizedmedianoperationsbRnd. Also,
\packageMinimizedpercentparentframesd and \packageMinimizedpercentparentframesb
of these expressions run in function environments.

\medskip\EE{$min(e)=\c{FUN}$} This form captures expressions that
define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs
in \packageFunctionDefinitionSitesPercent of sites. Evaluating a function
definition is done in 2 interpreter steps; the data does not record the work
performed by the interpreter when the generated functions are eventually run. In
addition to \c{FUN}, \packageGeneralizedFunctionDefinitionSitesPercent of sites
have function definitions nested in other expressions. This gives an idea of the
use of higher-order functions.

\medskip\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
code blocks, which occur in only 4\% of sites. These are huge expressions; we
do not inspect the contents of the blocks. The median number of executed operations
is \packageMinimizedmedianoperationsjRnd. They typically run in function
environments.

\medskip\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
returns a dataframe resulting from fitting the model described in a given
formula. This form subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the
single most popular function invoked from \eval; it occurs in 7\% of the sites.
Each call does quite a lot of work with a 2K instructions median.
